Type-Safe Modular Parsing

### 2017-01-13 paper/sections/ObjectAlgebras.tex

OO ASTs allow easily adding new language constructs, but it is hard
to add new operations over the AST. The OO AST in Figure~\ref{},
supports a pretty-printing operation. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

@@

\bruno{The following text is not very good, needs to be revised. I think
we need to show the full code and where it breaks. Since I removed text
about the Expression problem in S3, I think we need to talk more about it here.}
It refers to one Scala solution to the Expression Problem.\huang{what's the solution? does it have the later "critical issue"? if so, why is it a "solution"} But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

In contrast, Object Algebras~\cite{Oliveira2012} enables us to solve this problem. As a nice solution, it separates data variants
and operations, and offers high flexibility in the choice of operations to be performed over ASTs. Before adopting it for parsing, the next section introduces the background of Object Algebras.

@@

Note how the construction of the asbtract syntax happens throught
the use of factory methods, instead of constructors.
The pretty print the expression, resulting in \inlinecode{"(1 + x)"},
we pass an insteance of \inlinecode{VarExprPrint} to \inlinecode{makeExp}.

\begin{lstlisting}
makeExp(new VarExprPrint {})
\end{lstlisting}

\bruno{This is a good discussion, but it is out-of-place. We haven't
yet talking about parsing with object algebras, so I think this discussion
should come later.}
Now we observe that an ``object'' of expression actually
has the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{E} for generic \inlinecode{E}. Comparing with traditional objects, one may have concern that if a parser produces function structures, it could be hard to further process them. Especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform a \inlinecode{VarExprAlg} expression to an \inlinecode{ExprAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:\haoyuan{mention in discussion that shy deals with boilerplate.}

\lstinputlisting[linerange=45-56]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_REFACTOR
Here the \inlinecode{VarExprAlg} expression \inlinecode{"1 + x"} is changed to \inlinecode{ExprAlg} expression \inlinecode{"1 + 2"}, before the pretty-printing algebra is applied. The abstract method
\inlinecode{alg} in \inlinecode{Refactor} can actually be any algebra simply for delegation. By passing a concrete implementation
to \inlinecode{alg}, \inlinecode{Refactor} will be able to transform a \inlinecode{VarExprAlg} expression before we obtain the result.

>>>>

OO ASTs allow easily adding new language constructs, but it is hard
to add new operations over the AST. The code in Section~\ref{subsec:packratparsing},
supports a pretty-printing operation. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

@@

But then programmers have to define new classes for all the constructs, including literals, additions and variables, by extending the old ones together with \lstinline{FVarsExpr}, and implementing the corresponding \lstinline{fVars} respectively.

\begin{lstlisting}
class FVarsLit(x: Int) extends Lit(x) with FVarsExpr {
def fVars = Set()
}
class FVarsAdd(e1: FVarsExpr, e2: FVarsExpr) extends Add(e1, e2) with FVarsExpr {
def fVars = e1.fVars ++ e2.fVars
}
class FVarsVar(x: String) extends Var(x) with FVarsExpr {
def fVars = Set(x)
}
\end{lstlisting}
The above code already appears as boilerplate, especially, more classes have to be defined along with future extensions. It pollutes the namespace heavily. Moreover, the \lstinline{ExprParser} code is no longer useful; to generate expressions that support free variables, we have to substitute \lstinline{FVarsExpr}, \lstinline{FVarsLit}, \lstinline{FVarsAdd} and \lstinline{FVarsVar} for the old \lstinline{Expr}, \lstinline{Lit}, \lstinline{Add} and \lstinline{Var} correspondingly. Again, to avoid modifying existing code, we have to sacrifice code reuse. In fact, one would not like to make a choice from them, but expect both. We have to seek for a different approach to build ASTs. \haoyuan{Please check text.}

%\bruno{The following text is not very good, needs to be revised. I think
%we need to show the full code and where it breaks. Since I removed text
%about the Expression problem in S3, I think we need to talk more about it %here.}
%It refers to one Scala solution to the Expression Problem.\huang{what's the %solution? does it have the later "critical issue"? if so, why is it a %"solution"} But then programmers have to define new classes for literals, %additions and variables
%by extending the old ones together with \inlinecode{FreeVarsExpr}. The %critical issue is that, the existing parsing code only generates objects of %the old
%types, again it is a modularity issue, since it is a bad idea to modify %existing code again and again with extensions. Other solutions including %modelling constructors as functions introduce unnecessary complexity.

Fortunately, Object Algebras~\cite{Oliveira2012} enables us to solve this problem. As a nice solution, it separates data variants
and operations, and offers high flexibility in the choice of operations to be performed over ASTs. Before adopting it for parsing, the next section introduces the background of Object Algebras.

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty print the expression,
we pass an instance of \inlinecode{VarExprPrint} to \inlinecode{makeExp}:

\begin{lstlisting}
makeExp(new VarExprPrint {})
\end{lstlisting}
which results in \inlinecode{"(1 + x)"} as expected.
